---
title: "model development"
output: html_document
date: "2025-03-28"
---
Connect to database.

```{r}
library(tidyverse)
library(tidymodels)
library(RPostgreSQL)
library(httr)
library(jsonlite)
library(lubridate)

options(scipen = 999)

  dbname <- keyring::key_get("warehouse_name")
  host <- keyring::key_get("warehouse_host")
  port <- keyring::key_get("warehouse_port")
  user <- keyring::key_get("warehouse_user")
  password <- keyring::key_get("warehouse_password")

db <- RPostgreSQL::dbConnect(RPostgreSQL::PostgreSQL(),
                 dbname = dbname, 
                 host = host, 
                 port = port, 
                 user = user, 
                 password = password)
```

Query the customer class data from the data warehouse.

```{sql connection=db, output.var = 'class_table'}
SELECT c.utility_id, c.utility, c.report_year, c.customer_type, c.customer_count, c.mwh_delivered, c.revenue, u.utility_type
FROM energy.all_class c
LEFT JOIN energy.utility_info u ON c.utility_id = u.mn_utility_id 
/*WHERE report_year != 2023*/
```

Query county level sales from the data warehouse. This will represent the proportion of sales from a given county, not the total MWH sold to each county.

```{sql connection=db, output.var = 'county_sales'}
WITH total_mwh AS(
	SELECT utility_id, utility, report_year, mwh_delivered AS total_mwh
	FROM energy.all_class ac 
	WHERE customer_type = 'Total' /*AND report_year != 2023*/
)
SELECT c.utility_id, c.utility, c.report_year, c.county_name, (c.mwh_delivered / t.total_mwh) AS county_share
	FROM energy.county_sales c
	LEFT JOIN total_mwh t ON c.utility_id = t.utility_id AND c.utility = t.utility AND c.report_year = t.report_year
	/*WHERE c.report_year != 2023*/
```

Grab county fips codes from data warehouse

```{sql connection=db, output.var = county_fips}
SELECT co_name, co_code
FROM reference.county  
```

Query county level gdp figures from FRED. Multiply them by county sales proportions to estimate economic affects on electricity sales. 

```{r}
fips_list <- pull(county_fips, co_code)

fred_key <- keyring::key_get("fred_api_key")

fred_gdp_process <- function(fips){
  
  link <- paste0("https://api.stlouisfed.org/fred/series/observations?series_id=GDPALL", fips, "&api_key=", fred_key, "&file_type=json")
  
  raw_fred <- GET(link)
  
  fred_content <- fromJSON(rawToChar(raw_fred$content))
  
  fred_content <- fred_content$observations |>
    tibble() |> 
    mutate(fips_code = fips, 
           year = year(as.Date(date))) |> 
    inner_join(county_fips, by = c("fips_code" = "co_code")) |> 
    select(year, co_name, value) |> 
    rename(gdp = value)
  
  return(fred_content)
}

county_gdp_table <- fred_gdp_process(fips_list[1]) |> 
  slice(0)

for(i in fips_list){
  
  county_gdp_table <- fred_gdp_process(i) |> 
    bind_rows(county_gdp_table)
  
}

county_sales <- county_sales |> 
  left_join(county_gdp_table, by = c("report_year" = "year", "county_name" = "co_name")) |> 
  mutate(gdp_share = as.numeric(gdp) * county_share) |> 
  select(utility_id, utility, report_year, county_name, county_share, gdp_share)

```

Grab cooling and heating degree data by climate region from NOAA

```{r}
mn_climate_regions <- as.character(c(2101:2109))

cooling_links <- c(
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2024/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2023/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2022/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2021/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2020/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2019/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2018/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2017/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2016/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2015/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2014/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2013/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2012/ClimateDivisions.Cooling.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2011/ClimateDivisions.Cooling.txt"
)

link_cleaner <- function(link, year){
    
  x <- readLines(link)[145:153]
  n <- 1
  data_return <- tribble(~region_number, ~cooling_degree_days, ~year)
  for(i in x){
    region_data <- str_split(x[[n]][1], "\\|")[[1]]
    region_number <- region_data[1]
    cooling_degree_days <- sum(as.numeric(region_data[2:length(region_data)]))
    data_return <- tribble(~region_number, ~cooling_degree_days, 
                         region_number, cooling_degree_days) |> 
      mutate(year = year) |> 
      bind_rows(data_return) 
    n <- n + 1
  }
  
  data_return <- mutate(data_return, region_number = as.numeric(region_number))

  return(data_return)
}

cooling_data <- link_cleaner(cooling_links[1], 2024) |> 
  slice(0)

start_year <- 2024

for(i in cooling_links){
  
  cooling_data <- link_cleaner(i, start_year) |> 
    bind_rows(cooling_data)
  
  start_year <- start_year - 1
}
```

Heating Data
```{r}
heating_links <- c(
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2024/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2023/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2022/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2020/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2019/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2018/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2017/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2016/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2015/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2014/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2013/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2012/ClimateDivisions.Heating.txt", 
  "https://ftp.cpc.ncep.noaa.gov/htdocs/degree_days/weighted/daily_data/2011/ClimateDivisions.Heating.txt"
)

link_cleaner <- function(link, year){
    
  x <- readLines(link)[145:153]
  n <- 1
  data_return <- tribble(~region_number, ~heating_degree_days, ~year)
  for(i in x){
    region_data <- str_split(x[[n]][1], "\\|")[[1]]
    region_number <- region_data[1]
    heating_degree_days <- sum(as.numeric(region_data[2:length(region_data)]))
    data_return <- tribble(~region_number, ~heating_degree_days, 
                         region_number, heating_degree_days) |> 
      mutate(year = year) |> 
      bind_rows(data_return) 
    n <- n + 1
  }
  
  data_return <- mutate(data_return, region_number = as.numeric(region_number))
  return(data_return)
}


heating_data <- link_cleaner(heating_links[1], 2024) |> 
  slice(0)

start_year <- 2024

for(i in heating_links){
  
  heating_data <- link_cleaner(i, start_year) |> 
    bind_rows(heating_data)
  
  start_year <- start_year - 1
}
```

Combine county names with climate region numbers and apply county weights

```{r}
load(here::here("data/county_climate_regions.rda")) 
# county_climate_regions <- county_climate_regions |> 
#   left_join(heating_data, by = c("climate_region_number" = "region_number", "year" = "year")) |> 
#   left_join(cooling_data, by = c("climate_region_number" = "region_number", "year" = "year"))

county_sales <- county_sales |> 
  left_join(county_climate_regions, by = c("report_year" = "year", "county_name" = "county_name")) |> 
  mutate(heating_degree_days_weighted = county_share * heating_degree_days, 
         cooling_degree_days_weighted = county_share * cooling_degree_days) |> 
  select(utility_id, utility, report_year, county_name, county_share, gdp_share, heating_degree_days_weighted, 
         cooling_degree_days_weighted)
```


Combine data and pivot county columns out to function as predictors.

```{r}
county_vars <- county_sales |> 
  tibble() |> 
  pivot_wider(id_cols = c(utility_id, utility, report_year), names_from = county_name, values_from = county_share, names_prefix = "county_share_") |> 
  janitor::clean_names() |> 
  mutate(across(starts_with("county_share_"), .fns = ~replace_na(.x, 0)))
    #.cols = c(county_share_stevens:county_share_traverse), .fns = ~replace_na(.x, 0)))

gdp_vars <- county_sales |> 
  tibble() |> 
  pivot_wider(id_cols = c(utility_id, utility, report_year), names_from = county_name, values_from = gdp_share, names_prefix = "gdp_share_") |> 
  janitor::clean_names() |> 
  mutate(across(starts_with("gdp_share_"), .fns = ~replace_na(.x, 0)))

heating_vars <- county_sales |> 
  tibble() |> 
  pivot_wider(id_cols = c(utility_id, utility, report_year), names_from = county_name, values_from = heating_degree_days_weighted, names_prefix = "heating_days_") |> 
  janitor::clean_names() |> 
  mutate(across(starts_with("heating_days_"), .fns = ~replace_na(.x, 0)))

cooling_vars <- county_sales |> 
  tibble() |> 
  pivot_wider(id_cols = c(utility_id, utility, report_year), names_from = county_name, values_from = cooling_degree_days_weighted, names_prefix = "cooling_days_") |> 
  janitor::clean_names() |> 
  mutate(across(starts_with("cooling_days_"), .fns = ~replace_na(.x, 0)))


combined_data <- class_table |> 
  left_join(county_vars, by = c("utility_id" = "utility_id", "utility" = "utility", "report_year" = "report_year")) |> 
  left_join(gdp_vars, by = c("utility_id" = "utility_id", "utility" = "utility", "report_year" = "report_year")) |> 
  left_join(heating_vars, by = c("utility_id" = "utility_id", "utility" = "utility", "report_year" = "report_year")) |> 
  left_join(cooling_vars, by = c("utility_id" = "utility_id", "utility" = "utility", "report_year" = "report_year")) 
  tibble()

head(combined_data)

```

Check linearity of customer number and sales. Industrial customers do create some issues but generally the pattern is linear.

```{r}
class_table |> 
  filter(utility != 'Xcel Energy' & utility != 'Great River Energy') |> 
  ggplot() + aes(x = customer_count, y = mwh_delivered) + geom_point(aes(color = utility_type))
```

Plot revenue and sales. This is much more linear. 

```{r}
class_table |> 
  filter(utility != 'Xcel Energy' & utility != 'Great River Energy') |> 
  ggplot() + aes(x = revenue, y = mwh_delivered) + geom_point(aes(color = utility_type))
```

Split data into training and testing sets.

```{r}
set.seed(100)

combined_data_pre_23 <- combined_data |>
  tibble() |> 
  mutate(utility_type = as.factor(utility_type), 
         customer_type = as.factor(customer_type)) |> 
  filter(report_year != 2023) |> 
  filter(!c(utility %in% c("Xcel Energy", "Great River Energy")), 
         !c(customer_type %in% c("Total", "Irrigation", "Industrial")))

data_split <- initial_split(combined_data_pre_23, prop = 3/4)

train_data <- training(data_split)
test_data <- testing(data_split)
```

Specify recipe 

```{r}
class_rec <- recipe(
  mwh_delivered ~ ., data = train_data) |> 
  update_role(utility_id, utility, new_role = "ID") |> 
  update_role(report_year, new_role = "Year") |> 
  step_zv(all_predictors()) |> 
  step_unknown() |> 
  step_dummy(all_nominal_predictors()) |> 
  step_interact(terms = ~starts_with("customer_type"):customer_count) |> 
  step_interact(terms = ~starts_with("customer_type"):revenue) 
```

Specify models

```{r}
linear_mod <- linear_reg() |> 
  set_engine("lm")

```


Create workflows

```{r}
linear_workflow <- workflow() |> 
  add_model(linear_mod) |> 
  add_recipe(class_rec)

```

Fit linear model

```{r}
linear_mwh <- linear_workflow |> 
    fit(data = train_data)

```

Predict values in the test data

```{r}
mwh_results <- predict(linear_mwh, new_data = test_data) |> 
  bind_cols(predict(linear_mwh, new_data = test_data, type = "conf_int"))

```

Combine with actual results

```{r}
mwh_results_comb <- mwh_results |> 
  bind_cols(test_data |> select(utility, utility_id, mwh_delivered, utility_type))

```


RMSE
```{r}
results <- mwh_results_comb |> 
  select(.pred, mwh_delivered)

rmse(results, truth = mwh_delivered, estimate = .pred)$.estimate


```

Check Results with 2023 data
  
```{r}
data_23 <- combined_data |>
  filter(!c(utility %in% c("Xcel Energy", "Great River Energy")), 
         !c(customer_type %in% c("Total", "Irrigation", "Industrial"))) |> 
  mutate(utility_type = as.factor(utility_type), 
         customer_type = as.factor(customer_type)) |> 
  filter(report_year == 2023) 

```

Predict 2023 values with the model. 

```{r}
predictions_23 <- predict(linear_mwh, new_data = data_23) |> 
  bind_cols(predict(linear_mwh, new_data = data_23, type = "conf_int", level = .99)) |> 
  bind_cols(data_23 |> select(utility, utility_id, mwh_delivered, utility_type, customer_type, customer_count, revenue)) |>
  relocate(mwh_delivered, .after = .pred) 
```
Find outliers in 2023

```{r}
predictions_23 |> 
  filter(mwh_delivered > .pred_upper | mwh_delivered < .pred_lower) |> 
  mutate(maybe_unit_issue = case_when(
    abs(((mwh_delivered / 1000) / .pred) - 1) < .5 ~ TRUE, 
    TRUE ~ FALSE
  )) |> 
  #filter(.pred_upper > 0) |> 
  relocate(maybe_unit_issue, .before = utility) |> 
  arrange(utility)
```

```{r}
rmse(predictions_23, truth = mwh_delivered, estimate = .pred)
```



Linear model
```{r}
#saveRDS(linear_mwh, here::here("data/linear_mwh_model.rds"))
save(linear_mwh, file = here::here("data/linear_mwh_model.rda"))
```





